Detailed disease progression of 88 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis

### 2020-09-16 manuscript/manuscript.Rmd
Visualising multiverse

---
title: "Detailed disease progression of 88 patients hospitalized with Covid-19 in the Czech Republic: A dataset and exploratory analysis" 
output: pdf_document
abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to 10 hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. To assess association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom."
bibliography: Covid19.json
---

@@

The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of September 2020 led to over 28 million cases and over 900 000 deaths. Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteriods, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR; TODO]. Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteriod Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; http://zotero.org/users/5567156/items/5VAKEW58]. While initially promising, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM], as early treatment [@https://www.acpjournals.org/doi/10.7326/M20-4207] and as post-exposure profylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; http://zotero.org/users/5567156/items/GMR9FX9Z]. Further, no improvement could be discerned when adding Azithromycin to Hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ]. Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU]. No benefit was observed for convalscent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; http://zotero.org/users/5567156/items/BHGT88US] For additional summary see (live review)

In observational studies, Hydroxychloroquine was often found to be associated with better outcomes [e. g., @http://zotero.org/users/5567156/items/WZPNBRYN; http://zotero.org/users/5567156/items/4HR7MUJ6; http://zotero.org/users/5567156/items/NXC9CDI9]

@@

```{r}

@@

data_table_1 %>% group_by(group) %>% 
  summarise(
    N = n(),
    `Distinct sites` = length(unique(hospital_id)),
    Female = sum_percent(sex == "F"),
    `Age (mean, IQR)` = mean_iqr(age),
    `Admitted for Covid` = sum_percent(admitted_for_covid),
    `Took Azithromycin` = sum_percent(ever_az),
    `Ischemic Heart Disease` = sum_percent(ischemic_heart_disease),
    `Takes antihypertensives` = sum_percent(has_hypertension_drugs),	
    `Heart Failure` = sum_percent(heart_failure),
    COPD = sum_percent(COPD),
    Asthma = sum_percent(asthma),
    `Other lung disease` = sum_percent(other_lung_disease),
    Diabetes = sum_percent(diabetes),
    `Renal Disease` = sum_percent(renal_disease),
    `Liver Disease` = sum_percent(liver_disease),
    `Smoking` = sum_percent(smoking),
    `BMI (mean, IQR)` = mean_iqr(BMI),
    .groups = "drop"
  ) %>%
  mutate(group = factor(group, levels= c("All", "HCQ", "No HCQ", "Measured D-dimer","Measured IL-6"), ordered = TRUE)) %>%
  arrange(group) %>%
  as.data.frame() %>%
  column_to_rownames("group") %>%
  t() %>% 
  knitr::kable()
```

@@

```{r}
my_paste <- function(x) {
  paste0(as.integer(x), collapse = "")
}
wide  %>% 
  arrange(hospital_id, patient_id, day) %>%
  mutate(patient_id = factor(patient_id), patient_id = fct_reorder(patient_id, breathing,  .fun = my_paste, .desc = FALSE)) %>%
  ggplot(aes(x = day, y = patient_id, fill = breathing)) + 
  geom_tile() + 
  scale_fill_viridis_d(direction = -1) +
  theme(axis.text.y = element_blank()) +
  theme(axis.title.y = element_blank(), axis.ticks.y = element_blank())

@@

>>>>

---
title: "Detailed disease progression of 88 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  pdf_document
  # bookdown::pdf_document2: 
  #   toc: false
  ##bookdown::pdf_book: 
  #   base_format: rticles::arxiv_article
abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to 10 hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. To assess association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom."
bibliography: Covid19.json
---

@@

The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of September 2020 led to over 28 million cases and over 900 000 deaths. Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteriods, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR; TODO]. Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteriod Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58]. While initially promising, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI] and as post-exposure profylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z]. Further, no improvement could be discerned when adding Azithromycin to Hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ]. Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU]. No benefit was observed for convalscent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US] For additional summary see [@http://zotero.org/users/5567156/items/YP62P6XJ].

In observational studies, Hydroxychloroquine was often found to be associated with better outcomes [e. g., @http://zotero.org/users/5567156/items/WZPNBRYN; @http://zotero.org/users/5567156/items/4HR7MUJ6; @http://zotero.org/users/5567156/items/NXC9CDI9]

@@

```{r summarytable, echo=FALSE}

@@

data_table_1 %>% group_by(group) %>% 
  summarise(
    N = n(),
   # `Distinct sites` = length(unique(hospital_id)),
    Male = sum_percent(sex == "M"),
    `Age (mean, IQR)` = mean_iqr(age),
    `Admitted for Covid` = sum_percent(admitted_for_covid),
    `Took Azithromycin` = sum_percent(ever_az),
    `Ischemic Heart Disease` = sum_percent(ischemic_heart_disease),
    `Takes antihypertensives` = sum_percent(has_hypertension_drugs),	
    `Heart Failure` = sum_percent(heart_failure),
    COPD = sum_percent(COPD),
    Asthma = sum_percent(asthma),
    `Other lung disease` = sum_percent(other_lung_disease),
    Diabetes = sum_percent(diabetes),
    `Renal Disease` = sum_percent(renal_disease),
    `Liver Disease` = sum_percent(liver_disease),
    `Smoking` = sum_percent(smoking),
    `BMI (mean, IQR)` = mean_iqr(BMI),
    Deceased = sum_percent(outcome == "Death"),
    Discharged = sum_percent(outcome == "Discharged"),
    .groups = "drop"
  ) %>%
  mutate(group = factor(group, levels= c("All", "HCQ", "No HCQ", "Measured D-dimer","Measured IL-6"), ordered = TRUE)) %>%
  arrange(group) %>%
  as.data.frame() %>%
  column_to_rownames("group") %>%
  t() %>% 
  knitr::kable(caption = "Patient characterstics. Note that the 'Measured D-dimer' and 'Measured IL-6' are neither exclusive nor exhaustive.", booktabs = TRUE)
```

See Table \@ref(tab:summarytable) for the overall characteristics of the patient sample and several subgroups we used in the analysis.

@@

In Figure \@ref(fig:progression) we show the disease progression for all patients.

```{r progression, fig.cap="Disease progression for all patients included in the study"}
my_paste <- function(x) {
  paste0(as.integer(x), collapse = "")
}
wide  %>% 
  filter(!is.na(breathing)) %>%
  arrange(hospital_id, patient_id, day) %>%
  mutate(patient_id = factor(patient_id), patient_id = fct_reorder(patient_id, breathing,  .fun = my_paste, .desc = FALSE)) %>%
  ggplot(aes(x = day, y = patient_id, fill = breathing)) + 
  geom_tile() + 
  scale_fill_viridis_d(direction = -1) +
  theme(axis.text.y = element_blank()) +
  theme(axis.title.y = element_blank(), axis.ticks.y = element_blank())

@@

```{r multiverse_results}
bayesian_cols <- cols(
  hypothesis = col_character(),
  model = col_character(),
  adjusted = col_character(),
  estimand = col_character(),
  widest_CI_excl_reference = col_double(),
  point_estimate = col_double(),
  ci_low = col_double(),
  ci_high = col_double()
)
res_bayesian <- 
  rbind(read_csv(here::here("manuscript", "hmm_res.csv"), col_types = bayesian_cols),
        read_csv(here::here("manuscript", "jm_res.csv"), col_types = bayesian_cols)
        )

frequentist_cols <- cols(
  hypothesis = col_character(),
  model = col_character(),
  adjusted = col_character(),
  p_value = col_double(),
  ci_low = col_double(),
  ci_high = col_double()
)

res_frequentist <- 
  rbind(read_csv(here::here("manuscript", "competing_risks_res.csv"), col_types = frequentist_cols))
```

```{r, fig.height=6, fig.width=}
rbind(res_bayesian %>% select(-widest_CI_excl_reference), 
      res_frequentist %>% select(-p_value)) %>%
  mutate(
    model_adjusted = paste0(model, " - ", adjusted),
    label = factor(model_adjusted, levels = model_adjusted, labels = str_wrap(model_adjusted, 30)),
    hypothesis = factor(hypothesis, levels = names(hypotheses), labels = purrr::map_chr(hypotheses,.f = ~ .x$caption))) %>%
  ggplot(aes(x = point_estimate, xmin = ci_low, xmax = ci_high, y = label)) +
  geom_vline(xintercept = 0, color = "blue") +
  geom_point() +
  geom_errorbarh() +
  facet_wrap(~hypothesis, ncol = 1, scales = "free") 
  #theme(axis.text.y = element_text(size = 6))
```
